Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  partially-ordered \textbf{carrier set} \(P\) such that every \(p \in P\) has
  finitely many predecessors equipped with a mapping \(\alpha: P \to \mc{A}\)
  quotiented under the equivalence relation \(\alpha: P \to \mc{A} \simeq
  \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\) from \(P\)
  and \(Q\), there exists a partial order isomorphism between them which
  respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite} if its
  carrier set is. We will write unordered pomsets using multiset notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\), which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to \mc{A}\), where \(P + Q\) is given the lexicographic ordering and \(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many predecessors. They also form a monoid under parallel composition \(\alpha || \beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\) where \(P + Q\) is given the standard partial ordering (with elements of \(P\) and \(Q\) incomparable).

@@

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f || g = \lambda (a, b). 
  & \{\iota_0 ((a', b'), \alpha || \beta) 
    \mid \iota_0 (a', \alpha) \in f\;a 
    \land \iota_0 (b', \beta) \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
latter, but this definition does \textit{not} support sliding:
\begin{equation}
  (\ms{write}_x || \ms{id}) ; (\ms{id} || \ms{write}_y)
  \neq (\ms{id} || \ms{write}_y) ; (\ms{write}_x || \ms{id})
\end{equation}
since the left-hand side will emit the pomset \([x := a, y := b]\) while the right-hand side will emit the pomset \([y := b, x := a]\). If we did want this behaviour, we'd need to have a significantly more sophisticated model of composition which effectively takes nontermination into account, which we will leave to future work.

To graduate from SC concurrency to a genuine, if maximally simple, weak memory model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of lists of actions, which we will interpret as linear pomsets ordered by index.

\TODO{weak memory monad}

>>>>

We begin with the following definitions:
\begin{definition}[Pomset] 
  A \textbf{pomset} \(\alpha\) over a set of actions \(\mc{A}\) with a
  distinguished null action "\textbf{tick}" \(\delta \in \mc{A}\) is a
  \textit{nonempty} partially-ordered \textbf{carrier set} \(P\) such that every
  \(p \in P\) has finitely many predecessors equipped with a mapping \(\alpha: P
  \to \mc{A}\) quotiented under the equivalence relation \(\alpha: P \to \mc{A}
  \simeq \alpha': Q \to \mc{A}\) if, after removing finitely many \(\delta\)
  from \(P\) and \(Q\), there exists a partial order isomorphism between them
  which respects the values of \(\alpha, \alpha'\). A pomset is \textbf{finite}
  if its carrier set is. We will write unordered pomsets using multiset
  notation, e.g. \(\{a, a, b\}\), and linearly ordered pomsets using list
  notation, e.g. \([a, a, b]\).
\end{definition}
(Finite) pomsets form a monoid under sequential composition \(\alpha;\beta\),
which is defined by the function \([\alpha, \beta]: \ms{trim}(P + Q) \to
\mc{A}\), where \(P + Q\) is given the lexicographic ordering and
\(\ms{trim}(P)\) removes all elements of \(P\) with infinitely many
predecessors. They also form a monoid under parallel composition \(\alpha ||
\beta\), which is defined by the function \([\alpha, \beta]: P + Q \to \mc{A}\)
where \(P + Q\) is given the standard partial ordering (with elements of \(P\)
and \(Q\) incomparable). Note in both cases the monoidal unit is \(\{\delta\}\),
since we only consider nonempty pomsets but allow the removal of finitely many
ticks.

@@

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f || g = \lambda (a, b). 
  & \{\iota_0 ((a', b'), \alpha || \beta) 
    \mid \iota_0 (a', \alpha) \in f\;a 
    \land \iota_0 (b', \beta) \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
former, but this definition does \textit{not} support sliding:
\begin{equation}
  (\ms{write}_x || \ms{id}) ; (\ms{id} || \ms{write}_y)
  \neq (\ms{id} || \ms{write}_y) ; (\ms{write}_x || \ms{id})
\end{equation}
since the left-hand side will emit the pomset \([x := a, y := b]\) while the right-hand side will emit the pomset \([y := b, x := a]\). If we did want this behaviour, we'd need to have a significantly more sophisticated model of composition which effectively takes nontermination into account, which we will leave to future work.

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of
lists of actions, which we will interpret as linear pomsets ordered by index. In particular, we define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;\ms{PO}\), which gives us another \isotopessa model in the usual manner.
